---
title: "ICA8"
author: "Harrison Fringer"
date: "2023-04-04"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Loop Questions

Creating a loop that prints the nth letter of the alphabet in a specified format.
```{r}
for (o in 1:26){
  print(paste(letters[o], "is", "the", scales::ordinal(o), "letter", "of", "the", "alphabet."))
}
```
2. Using a count variable in a while loop that increases count by 1 while it is less than 40. It then checks if this iteration of count is divisible by 4.

```{r}
count <- 1
while(count < 40){
  count <- (count + 1)
  if (count%%4 == 0){
    next
  }
  print(count)
}
4%%2
```
## Vectorized Function Practice

3. Applying the summarize function to each level of the mtcars data set using apply()

```{r}
apply(X = mtcars, MARGIN = 2, FUN = summary) 
```
4. Using the Iris dataset and the tapply() function to find the mean and standard deviation for each Species of the Sepal.Length column.

```{r}
tapply(X = iris$Sepal.Length, INDEX = as.factor(iris$Species), FUN = mean)
tapply(X = iris$Sepal.Length, INDEX = as.factor(iris$Species), FUN = sd)
```
Then checking using the group by and summarize functions to prove they match.
```{r}
iris %>%
  group_by(Species) %>%
  summarize(mean = mean(Sepal.Length), sd = sd(Sepal.Length))
```

Running the test_vec through the chain of ifelse criteria, and printing the value.
```{r}
test_vec <- 100:150
test_vec <- ifelse(test_vec%% 15 == 0, "FizzBuzz", 
          ifelse(test_vec%% 5 == 0, "Buzz", ifelse(
            test_vec%% 2 == 0, "Fizz", test_vec)
          ))

print(test_vec)
```





